Generalising the impact of forest cover on streamflow from experimental data: it is not that simple.

### 2022-09-22 paper/Forest_and_Water/Forest_and_Water.tex
version for interbal review

@@

  \begin{abstract}
  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow. Using three different approaches, they all find a strong relationship between forestation, de-forestation and streamflow. However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation. The data for the recent three papers were reviewed, combined and re-analysed highlight the following: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow? Generalised additive models were used to run flexible regressions including multiple variables.
  Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow. Removal of forest cover causes a 32\% greater change in flow relative to increasing forest cover. Area of the catchment only affects the change in streamflow after log transformation, due to high skew in the data. Smaller catchment dominate the database with 42\% of the data \textless{} 1 km\textsuperscript{2} and 65\% of the data \textless{} 10 km\textsuperscript{2}. Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies. Despite these findings, overall explained variance (38\%) of the regression model is low due the quality of the inputs and additional unknown confounding factors.
  \end{abstract}
    \begin{keyword}
    keyword1 \sep 
    keyword2
  \end{keyword}

@@

The impacts of global deforestation and reforestation are important through their influence on streamflow and both blue and green water availability \citep{hoekvandijke2022, schyns2019}. The past work reviewing these impacts \citep{andreassian2004, jackson2005, zhang2017, brown2005, brown2013, filoso2017} highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa. The most dramatic result is Figure 5 in \citet{zhang2011} indicating (for Australian catchments) a 100\% decrease in streamflow for catchments with 100\% forest cover. However, on the other end of the spectrum, for three French catchments \citep{cosandey2005}, there was no change in streamflow characteristics in two of the catchments after deforestation. For reforestation, a modelling study across the 1 millon km\textsuperscript{2} Murray Darling Basin also found no major effect, especially in larger catchments \citep{vandijk2007}. Similarly a modelling study by \citet{beck2013} found no significant change in streamflows in 12 catchment in Puerto Rico as a result of deforestation. In contrast, in a recent study in Brazil across 324 catchments, \citet{levy2018} found a significant increase in streamflow, particular in the dry season, as a result of deforestation. This suggests that there can be significant variation across the different studies,methodologies and geographical regions.

@@

The most recent reviews \citep{zhang2017, filoso2017} developed an impressive global database of catchment studies with changes in streamflow due to changes in forest cover. The \citet{zhang2017} dataset, which covers over 312 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included. In contrast, the paper by \citet{filoso2017} focused primarily on reforestation, and covered an equally impressive database of 167 studies using a systematic review. In this case the collected data is mostly coded as count data and only a subset of 37 studies was analysed for actual water yield change. There is some overlap between the two data sets, but there are also some studies unique to both sets. The more regionally concentrated and detailed study by \citet{levy2018} is a further independent dataset with no overlap with the other studies. However, for this study only the flow and rainfall data is available for the catchments, and the change in landcover was derived from satellite data and was not made available.

@@

There are some clear limitations in these studies. The main method in the work by \citet{zhang2017} is a single covariate linear regression. In contrast, the systematic review from \citet{filoso2017} emphasises the classification and distributions of the study. \citet{zhang2017} points out that a main assumption in their work is that the catchment size threshold at 1000 km\textsuperscript{2} is a distinct separation between ``small'' and ``large'' catchments. However, a subset of 37 data points in \citet{filoso2017} (their Figure 9) does not appear to support this, suggesting a continuum. And while the work \citet{filoso2017} provides important insights in study types, analysis types, forest types and broad classification, there is limited quantification of actual impact.

@@

Given all these previous reviews and the seemingly clear conclusions about the impact of forest cover change on streamflow, the question is why another paper?
There is a real attraction in the idea of quantitative analysis of past studies to be able to extrapolate findings to larger scales and to identify factors across global scales.
However, there is also a real danger in this process, which is what we will highlight in this paper. There are four potential errors (or limitations) in the mentioned global analyses:

\begin{itemize}
\tightlist
\item
  Latent variables that are not included in the typical single covariate analysis;
\item
  Interpretation errors due to incomplete descriptions of the experiments in the original papers;\\
\item
  Aggregation of data that originates from different experiments with different objectives across a wide time period; and, finally\\
\item
  Transcription errors in the data, especially if data is collected from other review papers as some of the original papers are difficult to locate.
\end{itemize}

The aim of this paper is to highlight examples of each of these limitations, how they have influenced past work, and provide suggestions of how we can overcome these limitations. In addition, by applying more complex statistical models we will highlight the conclusions that can still be drawn from this work in relation to the impact of forest cover on streamflow. Finally, we will highlight future research needs in this area.

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3611}}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2083}}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4167}}@{}}
\caption{Summary of abbreviations of factors used in the Zhang et al.~(2017) data set}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\centering
Factor
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Abbreviation
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Definition
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\centering
Factor
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Abbreviation
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Definition
\end{minipage} \\
\midrule()
\endhead
forest type & CF & coniferous forest \\
& BF & broadleaf forest \\
& MF & mixed forest \\
hydrological regime & RD & rain dominated \\
& SD & snow dominated \\
climate type & EL & energy limited \\
& WL & water limited \\
& EQ & equitant \\
precipitation data type & OB & observed \\
& SG & spatial gridded \\
& MD & modelled \\
assessment technique & PWE & paired watershed experiment \\
& QPW & quasi-paired watershed
experiment \\
& HM & hydrological modelling \\
& EA & elasticity analysis \\
& SH & statistical modelling and
hydrographs \\
\bottomrule()
\end{longtable}

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4231}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1667}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1282}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}@{}}
\caption{\label{tab:m-all-linear} Statistical summary for the linear terms the full model}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule()
\endhead
\textbf{(Intercept)} & -7.47 & 15.87 & -0.47 & 0.64 \\
\textbf{DeltaF\_perc} & -0.59 & 0.05 & -10.89 & 0 \\
\textbf{Precip\_data\_typeOB} & -17.87 & 12.91 & -1.38 & 0.17 \\
\textbf{Precip\_data\_typeSG} & 0.2 & 14.82 & 0.01 & 0.99 \\
\textbf{Assessment\_techniqueEA, HM} & 18.67 & 41.7 & 0.45 & 0.65 \\
\textbf{Assessment\_techniqueHM} & 26.61 & 11.43 & 2.33 & 0.02 \\
\textbf{Assessment\_techniquePWE} & 30.77 & 11.68 & 2.63 & 0.01 \\
\textbf{Assessment\_techniquePWE,
HM} & 15.8 & 42.22 & 0.37 & 0.71 \\
\textbf{Assessment\_techniqueQPW} & 41.35 & 19.66 & 2.1 & 0.04 \\
\textbf{Assessment\_techniqueQPW,
EA} & 26.05 & 23.84 & 1.09 & 0.28 \\
\textbf{Assessment\_techniqueSH} & 39.31 & 11.54 & 3.41 & 0 \\
\textbf{Forest\_typeCF} & -9.28 & 7.41 & -1.25 & 0.21 \\
\textbf{Forest\_typeMF} & -6.34 & 7.38 & -0.86 & 0.39 \\
\textbf{Hydrological\_regimeSD} & 0.13 & 8.94 & 0.01 & 0.99 \\
\bottomrule()
\end{longtable}

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3472}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}@{}}
\caption{\label{tab:m-all-smooth} Statistical summary for the smooth terms for the full model}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
edf
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Ref.df
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
F
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
p-value
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
edf
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Ref.df
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
F
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
p-value
\end{minipage} \\
\midrule()
\endhead
\textbf{s(log10(Area\_km2))} & 0.77 & 4 & 0.85 & 0.03 \\
\textbf{s(Dryness)} & 4.71 & 9 & 2.18 & 0 \\
\textbf{s(length)} & 4.44 & 34 & 0.24 & 0.09 \\
\bottomrule()
\end{longtable}

The overall explaining power of the model can be interpreted from the adjusted \emph{r\textsuperscript{2}} (which is penalised for the number of parameters). This indicates an \emph{r\textsuperscript{2}} of 0.46 and deviance explained is 0.5, suggesting the model only explains about 50\% of the variance in the data.

@@

The overall partial slope of the change in forest cover is -0.59, if all other variables are kept constant. This suggest quite strong change in streamflow, moving from fully forested to fully cleared. Over the whole forest cover range, this is a change of -118 mm, with other variables held constant. This change is highly significant, as indicated by the low p-value.

In addition, all the smoothed variables \emph{log10(Area (km\textsuperscript{2}))} (p = 0.03)), \emph{Dryness} (p = 0)) and \emph{length} (p = 0.09)) explain variation in the data. For \emph{length}, the p-value is not strictly smaller, than 0.05, but still indicates some reasonable evidence that the variable explains some of the variation in the change in streamflow.

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4444}}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3056}}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.0972}}@{}}
\caption{\label{tab:tablecomparison} Comparison of alternative models for the relationship between the change in forest cover and the change in streamflow. (See Supplementary Material part 3)}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\centering
Model for change in forest
cover
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Deviation explained
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
AIC
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\centering
Model for change in forest
cover
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Deviation explained
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
AIC
\end{minipage} \\
\midrule()
\endhead
linear across range & 0.5 & 3167 \\
different for forestation and
deforestation & 0.46 & 3213 \\
non-linear across the range & 0.5 & 3167 \\
\bottomrule()
\end{longtable}

As discussed in the methods, the overall linear response to the change in forest cover was compared to a transformation of the negative forest cover to positives and a check whether the relationship might be non-linear. This approach tests whether the impact on streamflow from removing forest cover is different from reforestation, as outlined in the methods. The detail of the comparison is highlighted in Supplementary material part 3. However, generally the results of the analysis showed two main points (Table \ref{tab:tablecomparison}:

@@

As this is not always shown in papers discussing regression relationship, the residual distribution is provided in more detail (Figure \ref{fig:gamcheckmodelall}. Visually, the residuals appear approximately normal, although there is a noticable skew in a limited number of the data in the upper part of the distribution (Figure \ref{fig:gamcheckmodelall}). This is related to a limited number catchments that have very high changes in streamflow in the data set. In other words, the distribution of the residuals is somewhat fat-tailed.

@@

This model, which excludes data with long studies and very dry catchments explains only slightly less of the variation with an adjusted \emph{r\textsuperscript{2}} of 0.46 and a deviance explained of 0.49.

@@

One concern with the results presented so far is that there are a few assessment techniques in the data set with a very low number of observations and could influence the results of the analysis. This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM) (Table \ref{tab:tableassess}.

@@

As indicated before, if the overall variation due to this issues is small, then this would not be an issue for upscaling the results, but the large variation for the smaller catchments suggest that effects could be considerable. As \citet{jones2017} indicate, this really needs time series analysis of the different experiments. Some of the time series data might not be recoverable from the older experiments, which will limit the opportunities for analysis.

@@

\textbf{Maybe include a figure that shows the variation in the flow data by Assessment technique}

@@

\textbf{Still to do}

\begin{itemize}
\tightlist
\item
  studies are conditional on the data
\item
  residuals: should the data be further transformed\\
\item
  no interactions\\
\item
  implications for other ``meta-analysis'' studies\\
\item
  future research needs
\end{itemize}

>>>>

\usepackage{setspace}\doublespacing

@@

  \begin{abstract}
  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow. Using three different approaches, they all find a strong relationship between forestation, de-forestation and streamflow. The data for the recent three papers were reviewed, combined and re-analysed using generalised additive modelling. The overall results of the modelling indicates similar results to previous papers. The results are however problematic, the data set is unbalanced, and there are underlying correlations that suggest further factors that influence te results. The area of the catchment is strongly related to the assessment technique and the variability in the response data.
  A careful review of the data and the methodology in the previous papers highlights that there are four interlinked reasons that are problematic: 1) The existence of latent variables in the data that create the appearance of a relationship that really does not exist; 2) The difficulty in fully interpreting the specifics of different studies; 3) The difficulty of integrating data from seemingly similar studies, but with quite different objectives; and 4) The chance of transcription errors influencing the data. Overall this indicates that while valuable data can be extracted from past studies, the above problems need to be considered before results are extrapolated to greater and global scales.
  \end{abstract}
    \begin{keyword}
    data analysis \sep forest cover change \sep global scales \sep 
    literature review
  \end{keyword}

@@

There is a urgent need to identify the impacts of human intervention on streamflow at a global scale and to separate this from climate effects \citep{wang2020, hoekvandijke2022}. More specifically, the impacts of global deforestation and reforestation are important through their perceived influence on streamflow and blue and green water availability \citep{hoekvandijke2022, schyns2019}. The past work reviewing these impacts \citep{andreassian2004, jackson2005, zhang2017, brown2005, brown2013, filoso2017} highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa. The most dramatic result is Figure 5 in \citet{zhang2011} indicating (for Australian catchments) a 100\% decrease in streamflow for catchments with 100\% forest cover. However, on the other end of the spectrum, for three French catchments \citep{cosandey2005}, there was no change in streamflow characteristics in two of the catchments after deforestation. For reforestation, a modelling study across the 1 millon km\textsuperscript{2} Murray Darling Basin also found no major effect, especially in larger catchments \citep{vandijk2007}, but a recent study \citep{hoekvandijke2022} found an 8\% changes in streamflow as a result of reforestation. Similarly a modelling study by \citet{beck2013} found no significant change in streamflows in 12 catchment in Puerto Rico as a result of deforestation. In contrast, in a recent study in Brazil across 324 catchments, \citet{levy2018} found a significant increase in streamflow, particular in the dry season, as a result of deforestation. This suggests that there can be significant variation across the different studies,methodologies and geographical regions.

@@

There has been a recent push in the hydrological community \citep{evaristo2020metaanalysis} to use `meta-analysis' to summarise past studies. The claim is that because a meta-analysis uses clearly defined search terms and statistical methods to analyse the results, that this will lead to more reliable summaries of past research.

From the most recent reviews on forest impact on streamflow \citep{zhang2017, filoso2017}, the study by \citet{filoso2017} is a clear meta-analysis, while the other is not. However, both authors have developed an impressive global database of catchment studies with changes in streamflow due to changes in forest coverand use statistical approaches to analyse the resulting data. The \citet{zhang2017} dataset, which covers over 312 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included. In contrast, the paper by \citet{filoso2017} focused primarily on reforestation, and covered an equally impressive database of 167 studies using a systematic review. In this case the collected data is mostly coded as count data and only a subset of 37 studies was analysed for actual water yield change. There is some overlap between the two data sets, but there are also some studies unique to both sets. The more regionally concentrated and detailed study by \citet{levy2018} is a further independent dataset with no overlap with the other studies. However, for this study only the flow and rainfall data is available for the catchments, and the change in landcover was derived from satellite data and was not made available.

@@

There are some clear limitations in these studies, and some of this applies to meta-analyses in general. The main method in the work by \citet{zhang2017} is a single covariate linear regression. In contrast, the systematic review from \citet{filoso2017} emphasises the classification and distributions of the study. \citet{zhang2017} points out that a main assumption in their work is that the catchment size threshold at 1000 km\textsuperscript{2} is a distinct separation between ``small'' and ``large'' catchments. However, a subset of 37 data points in \citet{filoso2017} (their Figure 9) does not appear to support this, suggesting a continuum. And while the work \citet{filoso2017} provides important insights in study types, analysis types, forest types and broad classification, there is limited quantification of actual impact.

@@

Given all these previous reviews and the seemingly clear conclusions about the impact of forest cover change on streamflow, the question is why another paper?
There is a real attraction in the meta-analysis idea of statistical analysis of past studies to be able to extrapolate findings to larger scales and to identify factors across global scales \citep{evaristo2020metaanalysis}.
However, there is also a real danger in this process, which this paper aims to highlight. There are four potential errors (or limitations) in such global meta-analyses:

\begin{itemize}
\tightlist
\item
  Impact of latent variables that are not included in the typical single covariate analysis;
\item
  Interpretation errors due to incomplete descriptions of the experiments in the original papers;\\
\item
  Aggregation of data that originates from different experiments with different objectives across a wide time period, but have similar keywords; and, finally\\
\item
  Transcription errors in the data, especially if data is collected from other review papers as some of the original papers are difficult to locate.
\end{itemize}

The aim of this paper is to first reanalyse the global dataset using some more sophisticated statistical modelling and uses this to highlight examples of each of these limitation. This will show how they have influenced the outcomes of the past work, and provide suggestions of how we can overcome these limitations. In addition, by applying more complex statistical models we will highlight the conclusions that can still be drawn from this work in relation to the impact of forest cover on streamflow. Finally, we will highlight future research needs in this area.

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3611}}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2083}}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4167}}@{}}
\caption{\label{tab:table1} Summary of abbreviations of factors used in the Zhang et al.~(2017) data set}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\centering
Factor
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Abbreviation
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Definition
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\centering
Factor
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Abbreviation
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Definition
\end{minipage} \\
\midrule()
\endhead
forest type & CF & coniferous forest \\
& BF & broadleaf forest \\
& MF & mixed forest \\
hydrological regime & RD & rain dominated \\
& SD & snow dominated \\
climate type & EL & energy limited \\
& WL & water limited \\
& EQ & equitant \\
precipitation data type & OB & observed \\
& SG & spatial gridded \\
& MD & modelled \\
assessment technique & PWE & paired watershed experiment \\
& QPW & quasi-paired watershed
experiment \\
& HM & hydrological modelling \\
& EA & elasticity analysis \\
& SH & statistical modelling and
hydrographs \\
\bottomrule()
\end{longtable}

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4231}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1667}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1282}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}@{}}
\caption{\label{tab:m-all-linear} Statistical summary for the linear terms the full model}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule()
\endhead
\textbf{(Intercept)} & -4.81 & 16.31 & -0.29 & 0.77 \\
\textbf{DeltaF\_perc} & -0.6 & 0.06 & -10.71 & 0 \\
\textbf{Precip\_data\_typeOB} & -21.4 & 13.23 & -1.62 & 0.11 \\
\textbf{Precip\_data\_typeSG} & 9.36 & 15.17 & 0.62 & 0.54 \\
\textbf{Assessment\_techniqueEA, HM} & 20.64 & 42.73 & 0.48 & 0.63 \\
\textbf{Assessment\_techniqueHM} & 22.81 & 11.71 & 1.95 & 0.05 \\
\textbf{Assessment\_techniquePWE} & 30.63 & 11.94 & 2.57 & 0.01 \\
\textbf{Assessment\_techniquePWE,
HM} & 17.42 & 43.26 & 0.4 & 0.69 \\
\textbf{Assessment\_techniqueQPW} & 39.52 & 20.15 & 1.96 & 0.05 \\
\textbf{Assessment\_techniqueQPW,
EA} & 24.39 & 24.42 & 1 & 0.32 \\
\textbf{Assessment\_techniqueSH} & 45.3 & 11.83 & 3.83 & 0 \\
\textbf{Forest\_typeCF} & -9.45 & 7.6 & -1.24 & 0.21 \\
\textbf{Forest\_typeMF} & -8.05 & 7.56 & -1.06 & 0.29 \\
\textbf{Hydrological\_regimeSD} & 3.57 & 9.16 & 0.39 & 0.7 \\
\bottomrule()
\end{longtable}

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3472}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}@{}}
\caption{\label{tab:m-all-smooth} Statistical summary for the smooth terms for the full model}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
edf
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Ref.df
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
F
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
p-value
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
edf
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Ref.df
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
F
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
p-value
\end{minipage} \\
\midrule()
\endhead
\textbf{s(log10(Area\_km2))} & 0.79 & 4 & 0.99 & 0.02 \\
\textbf{s(Dryness)} & 4.64 & 9 & 2.26 & 0 \\
\textbf{s(length)} & 4.45 & 34 & 0.22 & 0.12 \\
\bottomrule()
\end{longtable}

The overall explaining power of the model can be interpreted from the adjusted \emph{r\textsuperscript{2}} (which is penalised for the number of parameters). This indicates an \emph{r\textsuperscript{2}} of 0.45 and deviance explained is 0.49, suggesting the model only explains about 50\% of the variance in the data.

@@

The overall partial slope of the change in forest cover is -0.6, if all other variables are kept constant. This suggest quite strong change in streamflow, moving from fully forested to fully cleared. Over the whole forest cover range, this is a change of -120 mm, with other variables held constant. This change is highly significant, as indicated by the low p-value.

In addition, all the smoothed variables \emph{log10(Area (km\textsuperscript{2}))} (p = 0.02)), \emph{Dryness} (p = 0)) and \emph{length} (p = 0.12)) explain variation in the data. For \emph{length}, the p-value is not strictly smaller, than 0.05, but still indicates some reasonable evidence that the variable explains some of the variation in the change in streamflow.

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4444}}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3056}}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.0972}}@{}}
\caption{\label{tab:tablecomparison} Comparison of alternative models for the relationship between the change in forest cover and the change in streamflow. (See Supplementary Material part 3)}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\centering
Model for change in forest
cover
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Deviation explained
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
AIC
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\centering
Model for change in forest
cover
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Deviation explained
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
AIC
\end{minipage} \\
\midrule()
\endhead
linear across range & 0.49 & 3182 \\
different for forestation and
deforestation & 0.45 & 3227 \\
non-linear across the range & 0.5 & 3182 \\
\bottomrule()
\end{longtable}

As discussed in the methods, the overall linear response to the change in forest cover was compared to a transformation of the negative forest cover to positives and a check whether the relationship might be non-linear. This approach tests whether the impact on streamflow from removing forest cover is different from reforestation, as outlined in the methods. The detail of the comparison is highlighted in Supplementary material part 3. However, generally the results of the analysis showed two main points (Table \ref{tab:tablecomparison}):

@@

As this is not always shown in papers discussing regression relationship, the residual distribution is provided in more detail (Figure \ref{fig:gamcheckmodelall}). Visually, the residuals appear approximately normal, although there is a noticable skew in a limited number of the data in the upper part of the distribution (Figure \ref{fig:gamcheckmodelall}). This is related to a limited number catchments that have very high changes in streamflow in the data set. In other words, the distribution of the residuals is somewhat fat-tailed.

@@

This model, which excludes data with long studies and very dry catchments explains only slightly less of the variation with an adjusted \emph{r\textsuperscript{2}} of 0.44 and a deviance explained of 0.48.

@@

One concern with the results presented so far is that there are a few assessment techniques in the data set with a very low number of observations and could influence the results of the analysis. This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM) (Table \ref{tab:tableassess}).

@@

As indicated before, if the overall variation due to this issues is small, then this would not be an issue for upscaling the results, but the large variation for the smaller catchments suggest that effects could be considerable. As \citet{jones2017} indicate, this really needs time series analysis of the different experiments. However, some of the time series data might not be recoverable from the older experiments, which will limit the opportunities for analysis. We will discuss this further below.

@@

@@

In this paper, a few studies have been singled out in the analysis. The choice of focus was mainly driven by the data that was made available by the authors of these papers \citep{zhang2017, filoso2017}, which provide a rich case study for the current paper.

Field research is by nature limited in space and time, due to the high costs involved of setting up experiments. This is particularly true for experiments in hydrology and forest hydrology, where field sites need to cover sufficient spatial and temporal variability. This means there is a general need to extrapolate the local results to larger scales to inform decision making and policy.

However, as is demonstrated in this paper, there are multiple issues when this local scale data is extrapolated to larger scales. It clearly demonstrates that the results of any model (in this case a regression model) is highly dependent on the data, but also on the assumptions in the model. From the perspective of extrapolating local data to global scales for policy advice and decision making \citep[i.e.][]{hoekvandijke2022, jackson2005}, this is an important point.

\hypertarget{residuals-of-the-model}{%
\subsubsection{Residuals of the model}\label{residuals-of-the-model}}

The residuals of the final model presented in this paper (Figure \ref{fig:gamcheckmodelall}) indicate that the residual distribution remains fat-tailed, causing deviations from an assumed \(\epsilon \sim N(0,\sigma^2)\). This once again highlights that there is unexplained variation at the extremes of the distribution, once again related to the paired watershed experiments (Figure \ref{fig:assessment}). Generally,in statistical models, the approach would be to further normalise the residuals through transformations. However, in this case this might be difficult and might not resolve all the issues due to the large variation in the data.

\hypertarget{interactions}{%
\subsubsection{Interactions}\label{interactions}}

The current modelling approach does not consider any interactions between the variables, and this would offer another approach to understand the variation in the data. As already indicated in Figure \ref{fig:corgraphs}, there are interactions between different variables. This further complicates the extrapolation of the local scale experiment data to global scales and to extend historical data to current management and decisions.

In this case, interactions were not included because, as was shown, there are bigger problems with trying to extrapolate the existing data, and the data itself can be problematic. To be able to model the interactions well, the nature of the variables and interactions need to be understood and or clearly hypothesized. Otherwise it becomes another case of correlation without causation.

\hypertarget{implications-for-other-meta-analysis-studies}{%
\subsubsection{Implications for other ``meta-analysis'' studies}\label{implications-for-other-meta-analysis-studies}}

There has been a recent push to develop more meta-analysis studies in hydrology \citep{wang2020, evaristo2020metaanalysis}, and we strongly believe that developing new insights by combining historical data sets from reviewed papers is highly valuable. However, this paper highlights that there is considerable chance that large historical data sets include latent variables and are more complex than envisioned. This is particularly true for more historical work, as methods of observation and even apporaches to management have changed considerably. The same management description is not necessarily the same action on the ground. A carefully designed and systematic approach can prevent some of bigger problems as is demonstrated in \citet{wang2020}, where both the approach and the catchment area are investigated as latent variables.
This is particularly relevant, where the results of meta-analyses are extrapolated to make global predictions without clearly quantified uncertainties (such as in \citet{hoekvandijke2022} and \citet{wang2020}).

A second potential danger is the extrapolation of the local small catchment results and conclusions tolarger scales, but beyond the original scope of the studies. For example, the current database is mainly related to forest harvest, bush fire and reforestation/plantation management. It is tempting to use the result of a large scale analysis of this data to make inferences about overall landuse change \citep{li2017, wang2020}, but this would not be valid, as the deforestation studies are generally not a transition to an agricultural landuse or pasture, but regrowing into forest. Similarly, using the plantation studies to extrapolate to ``reforestation'' (as in \citet{filoso2017} and \citet{hoekvandijke2022}) is also tenuous. Plantation forests are generally fast growing hybrids that will have quite different ecophysiology, particularly in South America \citep{jones2017, binkley2017}, while other reforestation, for example for salinity control in Australia, might focus on a mix of native species. Given the link between ecophysiology and water and carbon budgets, care should be taken in extrapolation, introducing a further error.
